Graded Call-By-Push-Value

### 2020-07-08 popl21/popl21.tex
[ popl21 ] semantics of graded CBPV

 \nru{\rvar^+}{}
     {\ValTy {\gqext 0\Gamma 1x P} x P}

@@

Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
$\qValTy \gamma \Gamma v P$ as families
$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
Naturality here means that
$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
    \cdot \gamma)}\right)_{r \leq s}$.
Again, because of $\rweak$, we interpret typing derivations rather than terms.
The interpretation is now rather straightforward, but we spell it out for reference.
\[
\begin{array}{llcl}
\rvar^+
  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
\\
\rintro\Box
  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose[\den \Gamma]
\\
\rintro\GS
  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
  & = & \iota_i \comp \denv v r
\\
\rintro\otimes
  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
\\
\end{array}
\]
\[
\begin{array}{llcl}
% \rvar^-
%   & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
%   & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
% \\
% \rlet^{-}
%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
%   & = & \dent t
%   \comp (\tid[\Den \Gamma \gamma] \otimes
%            (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
% \\ &&& {~}
%   \comp \duplicate
% \\
\rlet %^{+}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\Box
  & \dent {\qCompTy \gamma \Gamma {\force v} {N}}
  & = & \extract[\D] \comp \denv v 1
% \relim\Box
%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
%   & = & \dent t
%   \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
%   \comp \duplicate
\\
\relim\GS
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
  & = & [\dent {t_i}]_{i:I} \comp \tdist
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\otimes
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
  & = & \dent t \comp \isot
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\rintro\diamond
  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
  & = & \treturn \comp \denv v r
\\
\relim\diamond
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
\\ &&& {~}
  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
  \comp \duplicate
\\
\rintro\lolli
  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
  & = & \Lambda \dent t
\\
\relim\lolli
  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
\\
\rintro\Pi
  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
  & = & \langle \dent {t_i} \rangle_{i:I}
\\
\relim\Pi
  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
  & = & \pi_i \comp \dent t
\\
\rweak
  & \dent {\qCompTy \gamma \Gamma t N}
  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
\\
\end{array}
\]

@@

@@

\subsection{Coeffect-graded CBPV: substitution and metatheory}

@@

 \nru{\rvar^+}{}
     {\ValTy {\qext 0\Gamma 1x P} x P}

@@

      \gCompTy {\qext \gamma \Gamma r x P} t e N
    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} e N}

@@

      \forall i{:}I,\ \gCompTy {\qext \gamma \Gamma r {x_i} {P_i}} {t_i} e N
    }{\gpeCompTy {\gamma + r\delta} \Gamma {\caser v {\{r x_i.\, t_i\}_{i:I}}} e N}

@@

      \gqeCompTy \gamma {\extr \Gamma {\overline{r x_i{:}P_i}^{i:I}}} t e N
    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} e N}

@@

     {\gCompTy {\qext \gamma \Gamma r x P} t e N
    }{\gqeCompTy \gamma \Gamma {\lam x t} e {\qfun r P N}}

@@

     {\gqeCompTy \gamma \Gamma t e {\qfun r P N} \qquad

@@

    }{\gpeCompTy {\gamma + r\delta} \Gamma {\app t v} e N}

@@

     {\forall i{:}I,\ \gqeCompTy \gamma \Gamma {t_i} e {N_i}
    }{\gqeCompTy \gamma \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}

@@

     {\gqeCompTy \gamma \Gamma t e {\rect I N}
    }{\gqeCompTy \gamma \Gamma {\proj i t} e {N_i}}

@@

>>>>

 \nru{\rvar %^+
    }{
    }{\ValTy {\gqext 0\Gamma 1x P} x P}

@@

Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
$\qValTy \gamma \Gamma v P$ as families
$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
Naturality here means that
$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
    \cdot \gamma)}\right)_{r \leq s}$.
Again, because of $\rweak$, we interpret typing derivations rather than terms.
The interpretation is now rather straightforward, but we spell it out for reference.
\[
\begin{array}{llcl}
\rvar %^+
  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
\\
\rintro\Box
  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose[\den \Gamma]
\\
\rintro\GS
  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
  & = & \iota_i \comp \denv v r
\\
\rintro\otimes
  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
\\
\end{array}
\]
\[
\begin{array}{llcl}
% \rvar^-
%   & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
%   & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
% \\
% \rlet^{-}
%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
%   & = & \dent t
%   \comp (\tid[\Den \Gamma \gamma] \otimes
%            (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
% \\ &&& {~}
%   \comp \duplicate
% \\
\rlet %^{+}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\Box
  & \dent {\qCompTy \gamma \Gamma {\force v} {N}}
  & = & \extract[\D] \comp \denv v 1
% \relim\Box
%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
%   & = & \dent t
%   \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
%   \comp \duplicate
\\
\relim\GS
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
  & = & [\dent {t_i}]_{i:I} \comp \tdist
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\otimes
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
  & = & \dent t \comp \isot
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\rintro\diamond
  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
  & = & \treturn \comp \denv v r
\\
\relim\diamond
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
\\ &&& {~}
  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
  \comp \duplicate
\\
\rintro\lolli
  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
  & = & \Lambda \dent t
\\
\relim\lolli
  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
\\
\rintro\Pi
  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
  & = & \langle \dent {t_i} \rangle_{i:I}
\\
\relim\Pi
  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
  & = & \pi_i \comp \dent t
\\
\rweak
  & \dent {\qCompTy \gamma \Gamma t N}
  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
\\
\end{array}
\]

@@

The equational theory and operational semantics of coeffect-graded
CBPV is identical to the one of CBPV \citep[Fig.~11]{levy:hosc06}.  In
this section, we make just a few remarks on an alternative
presentation of the permutation laws (called ``sequencing laws'' in
\loccit).

@@

\subsection{Coeffect-graded CBPV: substitution and metatheory}
\label{sec:subst}

@@

 \nru{\rvar %^+
    }{
    }{\ValTy {\qext 0\Gamma 1x P} x P}

@@

      \CompTy {\qext \gamma \Gamma r x P} t e N
    }{\peCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} e N}

@@

      \forall i{:}I,\ \CompTy {\qext \gamma \Gamma r {x_i} {P_i}} {t_i} e N
    }{\peCompTy {\gamma + r\delta} \Gamma {\caser v {\{r x_i.\, t_i\}_{i:I}}} e N}

@@

      \qeCompTy \gamma {\extr \Gamma {\overline{r x_i{:}P_i}^{i:I}}} t e N
    }{\peCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} e N}

@@

     {\CompTy {\qext \gamma \Gamma r x P} t e N
    }{\qeCompTy \gamma \Gamma {\lam x t} e {\qfun r P N}}

@@

     {\qeCompTy \gamma \Gamma t e {\qfun r P N} \qquad

@@

    }{\peCompTy {\gamma + r\delta} \Gamma {\app t v} e N}

@@

     {\forall i{:}I,\
      \qeCompTy \gamma \Gamma {t_i} e {N_i}
    }{\qeCompTy \gamma \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}

@@

     {\qeCompTy \gamma \Gamma t e {\rect I N}
    }{\qeCompTy \gamma \Gamma {\proj i t} e {N_i}}

@@

\subsection{Graded CBPV: semantics}

As in effect-graded CBPV, negative types $N$ are interpreted as graded
monad algebras $\denn N : (\Eff,\leq) \to \CC$, and as in
coeffect-graded CBPV, positive types are interpreted as graded comonad
coalgebras $\denp P : (\R,\leq) \to \CC$.  Values
$\qValTy \gamma \Gamma v P$ are interpreted as natural transformations
$\dent v : \Denpar \Gamma {\_\cdot\gamma} \todot \denp P$ and
computations $\qeCompTy \gamma \Gamma t e N$ as morphisms
$\dent t : \Hom {\Den\Gamma \gamma} {\Denn N e}$.  Except for some
change in typing, the denotation of terms is unchanged from
coeffect-graded CBPV.  We recapitulate the most interesting cases here
with the updated typing:
\[
\xymatrix@C=10ex@R=3ex{
\denv {\thunk t} r
& \Den \Gamma {r\gamma}     \ar[r]^{\extract}
& \D[r]\,\Den\Gamma\gamma   \ar[r]^{\fmapt[{\D[r]}]{\dent t}}
& \D[r]\,\Denn N e
\\
\dent {\force v}
& \Den \Gamma \gamma    \ar[r]^{\denv v 1}
& \D[1]\,\Denn N e      \ar[r]^{\extract}
& \Denn N e
\\
\dent {\ret v}
& \Den \Gamma {r\gamma}  \ar[r]^{\denv v r}
& \Denp P r              \ar[r]^{\treturn}
& \T[\Ge]\,\Denp P r
\\
\dent {\bind x u t}
& \Den \Gamma {\gamma + \delta}                 \ar[r]^{\duplicate}
& \Den \Gamma \gamma \otimes \Den \Gamma \delta \ar[r]^{\tid \otimes \dent u}
& \Den \Gamma \gamma \otimes \T[e_1] \Denp P r  \ar[d]^{\strengthl}
\\
& \Denn N {e_1 \bu e_2}
& \T[e_1]\,\Denn N {e_2}                        \ar[l]_{\run[\denn N]}
& \T[e_1](\Den \Gamma \gamma \otimes \Denp P r) \ar[l]_{\fmapt[{\T[e_1]}] {\dent t}}
% \dent {\bind x u t}
% & \Den \Gamma {\gamma + \delta}                 \ar[r]^{\duplicate}
% & \Den \Gamma \gamma \otimes \Den \Gamma \delta \ar[r]^{\tid \otimes \dent u}
% & \Den \Gamma \gamma \otimes \T[e_1] \Denp P r  \ar[r]^{\strengthl}
% & \T[e_1](\Den \Gamma \gamma \otimes \Denp P r) \ar[r]^{\fmap {\dent t}}
% & \T[e_1]\,N_{e_2}                               \ar[r]^{\trun}
% & N_{e_1 \bu e_2}
}
\]
Subsumption (\rsub) is interpreted by pre- and post-composition with
the functorial actions of $\den \Gamma$ and $\denn N$:
\[
\xymatrix@C=10ex{
% \rsub
% &
\Den \Gamma {\gamma}   \ar[r]^{\Den\Gamma{\gamma \leq \gamma'}}
& \Den \Gamma {\gamma'}  \ar[r]^{\dent t}
& \Denn N e              \ar[r]^{\Denn N {e \leq e'}}
& \Denn N {e'}
}
\]

% \begin{align*}
% & \denv {\thunk t} r
% & &
% \xymatrix{
%   \Den \Gamma {r\gamma}     \ar[r]^{\extract}
% & \D[r]\,\Den\Gamma\gamma   \ar[r]^{\fmapt[{\D[r]}]{\dent t}}
% & \D[r]\,\Denn N e
% }
% \\
% \xymatrix{
%   \Den \Gamma \gamma    \ar[r]^{\denv v 1}
% & \D[1]\,\Denn N e      \ar[r]^{\extract}
% & \Denn N e
% }
% \\
% & \dent {\ret v}
% & &
% \xymatrix@C=8ex{
%   \Den \Gamma {r\gamma}  \ar[r]^{\denv v r}
% & \Denp P r              \ar[r]^{\treturn}
% & \T[\Ge]\,\Denp P r
% }
% \end{align*}

The operational semantics of graded CBPV is identical to the one of
CBPV, and the substitution typing of coeffect-graded CBPV
(Section~\ref{sec:subst}) can be mechanically transferred to fully
graded CBPV.
